Naturalness in U(1) extensions of the MSSM

### 2014-12-04 paper/e6ssm-tuning-paper.tex
Added partial derivatives needed for master formula
Changed phrasing in some sentences

The sensitivity coefficients $\Delta_p$ may be calculated directly from the expression for $M_Z^2$ in terms of the $p_i$ for a particular model, which leads to a so-called master formula for calculating the fine tuning. A master formula for the E$_6$SSM, obtained from the tree level scalar potential, was presented in \cite{Athron:2013ipa}. In order to derive the expression presented there, the fact that $s \gg v$ was made use of to neglect certain $\mathcal{O}(v^2)$ terms in the EWSB conditions, greatly simplifying the final result.
% Important question:
% Impact of varying charges, is it picked up by the approximate CE6SSM master formula?
% e.g. how different are things in the limit Q_S --> 0?
For the purposes of exploring a wider class of $E_6$ inspired models, we have derived the master formula without neglecting any $\mathcal{O}(v^2)$ terms. The more complete tree-level master formula is somewhat complicated. This is because, unlike in the MSSM, even at tree level it is not possible to solve explicitly for the VEVs $v_1$, $v_2$ in terms of the Lagrangian parameters. It may be written in the form

@@

where the sum is over all low energy running parameters appearing in the tree level EWSB conditions, i.e. $q\in\{\lambda, A_\lambda, m_{H_d}^2,m_{H_u}^2,m_S^2,g_1,g_2,g_1'\}$. Explicit expressions for the function $C$ and the $\tilde{\Delta}_q$ appearing above are given in Appendix \ref{app:masterformula}. It should be noted that the effects of $U(1)$ mixing are neglected in deriving Eq. (\ref{eq:E6masterformula}).\\ \\
Evaluating the fine tuning directly using a master formula approach rapidly becomes impractical, as a brief inspection of the expressions for the $\tilde{\Delta}_i$ should make clear. It is well known in the MSSM that radiative corrections can significantly change (indeed, reduce) the fine tuning \cite{Cassel:2010px}. It is therefore important when studying the fine tuning to include 1- and 2-loop corrections in the fine tuning measure. To do so it is most convenient to work in terms of the EWSB conditions Eq. (\ref{eq:E6EWSBConditions}), rather than Eq. (\ref{eq:E6MZequation}). The general procedure is as follows \cite{Ellwanger:2011mu}. For a model in which $m$ fields develop real VEVs (e.g. $m=2$ in the MSSM, $m=3$ in the NMSSM and in the $E_6$ models considered), we require that the $m$ minimization conditions

@@

The most involved step in the process of getting $\Delta_p$ is evaluating the matrix containing the derivatives $\partial q_k/\partial p$. In general this must be done numerically, by evaluating the running using the 1- and 2-loop RGEs. This is an obstacle to doing large scans of the parameter space, as the process of numerically evaluating $n\times l$ derivatives is computationally expensive. In general, for example for models defined at the GUT scale, accurately calculating the fine tuning is better done by numerically differentiating $M_Z^2$ directly to reduce the number of derivatives that must be computed. For studying models defined at low energies, as we do here, we can take advantage of the fact that the amount of running involved is small. This makes it possible to use approximate analytic solutions to the RGEs that exhibit good accuracy over the range of scales considered. Given the 2-loop RG equation for a parameter $q$,

@@

\begin{eqnarray}
z_1 &=& \frac{\partial f_2}{\partial s}\frac{\partial f_3}{\partial \tan\beta}-\frac{\partial f_2}{\partial \tan\beta}\frac{\partial f_3}{\partial s}\nonumber\\
z_2 &=& \frac{\partial f_3}{\partial s}\frac{\partial f_1}{\partial \tan\beta}-\frac{\partial f_3}{\partial \tan\beta}\frac{\partial f_1}{\partial s}\nonumber\\
z_3 &=& \frac{\partial f_1}{\partial s}\frac{\partial f_2}{\partial \tan\beta}-\frac{\partial f_1}{\partial \tan\beta}\frac{\partial f_2}{\partial s}\nonumber,
\end{eqnarray}
with $f_1,f_2,f_3$ as given in Eq. (\ref{eq:E6EWSBConditions}). Explicitly, we find that
\begin{eqnarray*}
z_1&=&%\frac{M_Z\cos^2\beta}{2\bar{g}^3}\Big [ 4M_Z^2\Big ( \sqrt{2}\lambda A_\lambda \cos\beta-2s(\lambda^2+g_1'^2\tilde{Q}_2\tild%e{Q}_S )\sin\beta\Big )\nonumber\\
%&& \times \Big ( \sqrt{2}\lambda A_\lambda \cos 2\beta+g_1'^2(\tilde{Q}_1-\tilde{Q}_2)\tilde{Q}_Ss\sin 2\beta \Big )\Big ]
\end{eqnarray*}
For a running parameter $q$ appearing in the tree level EWSB conditions, the corresponding contribution to the sensitivity coefficient may be written 

@@

\frac{\partial f_2}{\partial M_Z}&=&\frac{2}{\bar{g}}\,\sin\beta\left ( m_{H_u^2}+\frac{\lambda^2s^2}{2}+\frac{g_1'^2}{2}\tilde{Q}_2\tilde{Q}_Ss^2+\frac{6g_1'^2}{\bar{g}^2}\tilde{Q}_2^2M_Z^2\right )-\sqrt{2}\frac{\lambda A_\lambda s}{\bar{g}}\,\cos\beta \nonumber\\

>>>>

The sensitivity coefficients $\Delta_p$ may be calculated directly from the expression for $M_Z^2$ in terms of the $p_i$ for a particular model, which leads to a so-called master formula for calculating the fine tuning. A master formula for the E$_6$SSM, obtained from the tree level scalar potential, was presented in \cite{Athron:2013ipa}. In order to derive the expression presented there, the fact that $s \gg v$ was made use of to neglect certain $O(v^2)$ terms in the EWSB conditions, greatly simplifying the final result. For the purposes of exploring a wider class of $E_6$ inspired models, we have rederived the master formula without neglecting any $O(v^2)$ terms. The more complete tree-level master formula is somewhat complicated. This is because, unlike in the MSSM, even at tree level it is not possible to solve explicitly for the VEVs $v_1$, $v_2$ in terms of the Lagrangian parameters. It may be written in the form

@@

where the sum is over all low energy running parameters appearing in the tree level EWSB conditions, i.e. $q\in\{\lambda, A_\lambda, m_{H_d}^2,m_{H_u}^2,m_S^2,g_1,g_2,g_1'\}$. Expressions for the quantity $C$ and the $\tilde{\Delta}_q$ appearing above are given in Appendix \ref{app:masterformula}. It should be noted that the effects of $U(1)$ mixing are neglected in deriving Eq. (\ref{eq:E6masterformula}).\\ \\
However, it is well known in the MSSM that radiative corrections can significantly change (indeed, reduce) the fine tuning \cite{Cassel:2010px}. It is therefore important when studying the fine tuning to include 1- and 2-loop corrections to the effective potential in the fine tuning measure. To do so it is most convenient to work in terms of the EWSB conditions Eq. (\ref{eq:E6EWSBConditions}), rather than Eq. (\ref{eq:E6MZequation}). The general procedure is as follows \cite{Ellwanger:2011mu}. For a model in which $m$ fields develop real VEVs (e.g. $m=2$ in the MSSM, $m=3$ in the NMSSM and in the $E_6$ models considered), we require that the $m$ minimization conditions

@@

Evaluating the derivatives $\partial q_k/\partial p$ must (in general) be done numerically, by numerically integrating the 2-loop RGEs. This is time consuming and presents an obstacle to doing large scans of the parameter space. For studying models defined at low energies, as we do here, we can take advantage of the fact that the amount of running involved is small. This makes it possible to use approximate analytic solutions to the RGEs that exhibit good accuracy over the range of scales considered. Given the 2-loop RG equation for a parameter $q$,

@@

\begin{equation}
z_i =\epsilon_{ijk} \frac{\partial f_j}{\partial s}\frac{\partial f_k}{\partial \tan\beta}
\label{eq:masterformulacoefficients}
with $f_1,f_2,f_3$ as given in Eq. (\ref{eq:E6EWSBConditions}). The relevant partial derivatives are
\frac{\partial f_1}{\partial \tan\beta}&=&-\frac{2M_Z}{\bar{g}}\,\cos^2\beta\left \{ \frac{\lambda A_\lambda s}{\sqrt{2}}\,\cos\beta+\sin\beta \left [ m_{H_d}^2+\frac{s^2}{2}\left ( \lambda^2+g_1'^2\tilde{Q}_1\tilde{Q}_S \right )\right.\right.\nonumber\\
&&\left.\left.+M_Z^2\left ( \frac{5}{2}-\frac{4\lambda^2}{\bar{g}^2}-\frac{4g_1'^2}{\bar{g}^2}\tilde{Q}_1\tilde{Q}_2+\frac{6g_1'^2}{\bar{g}^2}\tilde{Q}_1^2\right )\right ]+3M_Z^2\,\sin^3\beta\left [ \frac{2\lambda^2}{\bar{g}^2}-1\right.\right.\nonumber\\
&&\left.\left.+\frac{2g_1'^2}{\bar{g}^2}\left ( \tilde{Q}_1\tilde{Q}_2-\tilde{Q}_1^2\right )\right ]\right \},\nonumber\\
\frac{\partial f_1}{\partial s}&=&\frac{2M_Z}{\bar{g}}\left [ s\left ( \lambda^2+g_1'^2\tilde{Q}_1\tilde{Q}_S\right )\,\cos\beta-\frac{\lambda A_\lambda}{\sqrt{2}}\,\sin\beta \right ],\nonumber\\
\frac{\partial f_2}{\partial \tan\beta}&=&\frac{2M_Z}{\bar{g}}\,\cos^2\beta\left \{ \frac{\lambda A_\lambda s}{\sqrt{2}}\,\sin\beta+\cos\beta\left [ m_{H_u}^2+\frac{s^2}{2}\left ( \lambda^2+g_1'^2\tilde{Q}_2\tilde{Q}_S \right )\right.\right.\nonumber\\
&&\left.+M_Z^2\left ( \frac{5}{2}-\frac{4\lambda^2}{\bar{g}^2}-\frac{4g_1'^2}{\bar{g}^2}\tilde{Q}_1\tilde{Q}_2+\frac{6g_1'^2}{\bar{g}^2}\tilde{Q}_2^2\right )\right ]+3M_Z^2\,\cos^3\beta \left [ \frac{2\lambda^2}{\bar{g}^2}-1\right.\nonumber\\
&&\left.\left.+\frac{2g_1'^2}{\bar{g}^2}\left ( \tilde{Q}_1\tilde{Q}_2-\tilde{Q}_2^2\right )\right ]\right \},\nonumber\\
\frac{\partial f_2}{\partial s}&=&\frac{2M_Z}{\bar{g}}\left [ s\left ( \lambda^2+g_1'^2\tilde{Q}_2\tilde{Q}_S\right )\,\sin\beta-\frac{\lambda A_\lambda}{\sqrt{2}}\,\cos\beta\right ],\nonumber\\
\frac{\partial f_3}{\partial \tan\beta}&=&\frac{2M_Z^2}{\bar{g}^2}\,\cos^2\beta\left [ g_1'^2\tilde{Q}_Ss\left ( \tilde{Q}_2-\tilde{Q}_1\right )\,\sin 2\beta-\sqrt{2}\lambda A_\lambda\,\cos 2\beta\right ],\nonumber\\
\frac{\partial f_3}{\partial s}&=&m_S^2+\frac{2\lambda^2M_Z^2}{\bar{g}^2}+\frac{g_1'^2}{2}\tilde{Q}_S\left [\frac{4M_Z^2}{\bar{g}^2}\left ( \tilde{Q}_1\,\cos^2\beta+\tilde{Q}_2\,\sin^2\beta\right ) +3\tilde{Q}_S s^2\right ].\nonumber
For a running parameter $q$ appearing in the tree level EWSB conditions, the corresponding contribution to the sensitivity coefficient can then be written 

@@

\frac{\partial f_2}{\partial M_Z}&=&\frac{2}{\bar{g}}\,\sin\beta\left ( m_{H_u}^2+\frac{\lambda^2s^2}{2}+\frac{g_1'^2}{2}\tilde{Q}_2\tilde{Q}_Ss^2+\frac{6g_1'^2}{\bar{g}^2}\tilde{Q}_2^2M_Z^2\right )-\sqrt{2}\frac{\lambda A_\lambda s}{\bar{g}}\,\cos\beta \nonumber\\
